---
title:   "Interactive plot with trelliscopejs"
site:    workflowr::wflow_site
author:  "Author: Peter Pis&aacute;r"
date:    "`r paste(\"Last update:\", format(lubridate::with_tz(as.POSIXct(Sys.time()) + 7200, tzone = \"GMT\"), \"%Y-%m-%d %H:%M GMT+2\"))`"
purpose: interactive scatter plot with qualitative colorscales
output:
  workflowr::wflow_html:
    code_folding: hide
editor_options:
  chunk_output_type: console
# note:  -
---

```{r, include = F}
knitr::opts_chunk$set(
  comment = "#>",
  eval = T
)
```


# Interactive scatter plots with regression lines
These scatter plots are created also by using `trelliscopejs::facet_trelliscope()`.

One of advantages is a possibility to manage displaying plots using the left pane.

One of disadvantages is a warning message that pops-up after using `as_plotly = TRUE`.

  - More about this warning is [here](https://github.com/hafen/trelliscopejs/issues/61).
  - It is also reflected in a created HTML file in a result text informing that some file
  couldn't be loaded. JavaScript is used in the source .Rmd file to hide this unwanted message.

More about using package `trelliscopejs` is [here](https://github.com/hafen/trelliscopejs) and
in [README](https://github.com/LearnUseZone/rpkgs/tree/master/code/workflowrsubdirs) file.


# Load prepared data
```{r trelliscopejs_src, class.source = 'fold-show'}
knitr::read_chunk("code/examples/iris/interactive-visuals/trelliscopejs.R")
```


# Scatter plots
```{r trelliscopejs_Rmd_input, eval = T, warning = F, message = F}
```


```{js hide_text_err_result, echo = F}
function hide_result_text() {
  var subelements = document.getElementById("scatter-plots").childNodes;
  subelements[5].style.display = "none";
}
window.onload = hide_result_text()
```

