---
title:   "Data overview of iris dataset"
site:    workflowr::wflow_site
author:  "Author: learnusezone@gmail.com"
date:    "`r paste(\"Last update:\", format(lubridate::with_tz(as.POSIXct(Sys.time()) + 7200, tzone = \"GMT\"), \"%Y-%m-%d %H:%M GMT+2\"))`"
purpose: basic data overview of iris dataset
output:
  workflowr::wflow_html:
    toc: true
    highlight: pygments
    theme: paper
    code_folding: hide
editor_options:
  chunk_output_type: console
# from-to: 2020-12-09-
# note:  -
---

```{r, include = F}
knitr::opts_chunk$set(
  # base::setwd(here::here()),  # unnecessary because it's a part of workflowrsubdirs::build_htmls()
  comment = "#>",
  eval = T
)
```


# History of iris dataset
> The Iris flower data set or Fisher’s Iris data set is a multivariate data set introduced by
the British statistician and biologist Ronald Fisher in his 1936 paper. \
The use of multiple measurements in taxonomic problems as an example of linear discriminant analysis. \
The data set consists of 50 samples from each of three species of
Iris (Iris setosa, Iris virginica and Iris versicolor). \
Four features were measured from each sample: the length and
the width of the sepals and petals, in centimeters.
> `r tufte::quote_footer('--- source: https://rpubs.com/AjinkyaUC/Iris_DataSet')`


# Load prepared data
```{r data_overview, class.source = 'fold-show'}
source("code/examples/iris/data-overview.R")
```


# Data overview of iris dataset
## Observations, variables, names
Dataset `iris` has **`r dimensions[1]` observations**
and **`r dimensions[2]` variables**
and its **class name** is **\"`r class_name`\"**.

```{r names, comment = ''}
base::message("Variable (column) names of iris dataset are:", "\n",
  base::paste(var_names, collapse = "\n")          
)
```

## Head, tail, summary
```{r head-and-tail, class.source = 'fold-show'}
# the 1st 4 and the last 4 observations
base::print(base::rbind(head_iris, tail_iris))

# print summary
base::print(summary_iris)
```
