---
title:  "[workflowrsubdirs](https://github.com/LearnUseZone/workflowrSubfolders)"
author: LearnUseZone
date:   "`r paste('Last update:', format(lubridate::with_tz(as.POSIXct(Sys.time()) + 7200, tzone = 'GMT'), '%Y-%m-%d %H:%M GMT+2'))`"
output:
  github_document:
    toc: true
    toc_depth: 1
    html_preview: true
---

```{r, include = FALSE}
knitr::opts_chunk$set(eval = F)
```


## Purpose
This package is used (only) to build HTML files
from their associated R Markdown files saved in a new directory and its subdirectories
of a [workflowr](https://github.com/jdblischak/workflowr) project.

- [Package workflowr offers](https://github.com/jdblischak/workflowr/blob/master/README.md)
many useful features but doesn\'t offer this feature of package `workflowrsubdirs`.



## General rules
### Directory structure
- Because this package is a kind of an (optional) extension for package `workflowr`,
  it's necessary to adhere the structure of the main
  [workflowr project\'s working directory](https://jdblischak.github.io/workflowr/articles/wflow-01-getting-started.html)
  (hereinafter `main workflowr dir`). The most important directories to follow this rule are:
  - `analysis`
  - `code`
  - `docs` (for GitHub) or `public` (for GitLab)


### Directories for R Markdown files
- 1stly, `workflowr` project needs to be prepared.
    - Relevant steps (together with more information) can be found
      [here](https://jdblischak.github.io/workflowr/articles/wflow-01-getting-started.html).
      Read e.g. about directories:
      - `code` that can contain also subdirectories.
      - `docs` that contain all HTML files for your website.
- 2ndly, ensure that `docs/index.html` is created, e.g by running `workflowr::wflow_build()`.
    - Ideal is to also properly prepare files `_site.yml` and `index.Rmd` in directory `analysis`
      and also directories with HTML files included in `_site.yml` (e.g. `footer.html`).
- For the purpose of saving R Markdown files in subdirectories, it's necessary
  to create a new directory directly in `main workflowr dir`.
    - Don't create subdirectories (for this purpose) in directory `analysis`.
    - A required name of this new directory is `code-rmd`.
      - R Markdown files saved in it will be usually associated to solutions in directory `code`.
- R Markdown files can be saved
    - directly in this new directory or in any of its subdirectories (hereinafter `code-rmd subdirs`).
- Following 2 approaches can be used to show results of solutions from directory `code`
  and keep them organized:
    1. Create associated R Markdown files to directory `analysis` and
       build their HTML files using package `workflowr`.
       - this approach cannot have the same structure as subdirectories in directory `code`.
    2. Create associated R Markdown files to `code-rmd subdirs` and build
       their HTML files using package `workflowrsubdirs`.
       - This approach can have the same structure as directory \"code\".


### Open workflowr project before using workflowrsubdirs
- If other than `workflowr` project is originally opened (using its .Rproj file)
  then function `workflowrsubdirs::build_htmls()` fails because
  `base::setwd(here::here())` in this function sets a working directory to
  an original .Rproj working directory regardless a (later changed) current working directory.


### Only files with extension .rmd or .Rmd are allowed
It\'s required that original (source) R Markdown files have an extension .rmd or .Rmd. \
This behavior is in accordance with package `workflowr` where if the file extension is
not .rmd or .Rmd then rendering will fail with error \"Error: Invalid input for argument files\".



## Package functions
### build_htmls()
This is currently the only one function accessible by users.

Main steps:

  1. Evaluate if rendering of R Markdown to HTML files is possible. If some check doesn\'t pass,
     write a reason and stop processing.
     - Used package function: initial_checks()
  2. Create paths to original R Markdown files. Non-existent R Markdown files are ignored.
     An R Markdown file referred by more than 1 regular expression is processed only once.
     - Used package function: create_rmd_paths()
  3. Create a temporary copy of original R Markdown files to directory "code-rmd"
     (it's needed to be able to use figures generated by functions like e.g. graphics::hist()).
     Generate temporary (helping) R Markdown files from associated
     temporary code-rmd R Markdown files and save them to directory `analysis`.
     - Used package function: render_to_htmls()
  4. If input parameter `commit == TRUE`, create a commit of these
     temporary analysis R Markdown files with text
     \"feat: Commit temporary analysis .(r|R)md files\".
     - Used function: workflowr::wflow_git_commit()
     - If these temporary analysis R Markdown files are not committed before
       workflowr::wflow_build(), the following error message is written after
       opening any of prepared HTML files and
       left-clicking on `workflowr` button under tab \"Checks\":
       \"R Markdown file: uncommitted changes\"
         - This is also inline with `workflowr` package.
  5. Render temporary (helping) R Markdown files in directory `analysis` to HTML files.
     - Used function: workflowr::wflow_build()
     - Final HTML files are prepared (also package `workflowr` is set in this way)
       in directory `docs` (for GitHub) or directory `public` (for GitLab).
       Each such file name contains `\-\-` which are delimiters
       for paths to original R Markdown file paths.
  6. Delete temporary R Markdown files in directories `code-rmd` and `analysis`.



## Installation steps (for Windows)
1. Clone this Git [rpkgs repository](https://github.com/LearnUseZone/rpkgs).
2. Open `rpkgs/code/workflowrsubdirs/workflowrsubdirs/workflowrsubdirs.Rproj` in RStudio.
3. Build (from RStudio top menu) -> Build Source Package -> wait until .tar.gz file is created.
4. Unload package `workflowrsubdirs` (if it's loaded within tab \"Packages\") to avoid
   the following warning: \
      Warning in install.packages : \
      \ \ package ‘workflowrsubdirs’ is in use and will not be installed
   - Examples: \
     4a) Session (from RStudio top menu) -> Restart R (Ctrl+Shift+F10) \
     4b) Open tab \"Packages\" -> find package `workflowrsubdirs` -> uncheck a relevant check box \
     4c) `detach("package:workflowrsubdirs", unload = TRUE)`
5. Install the package using R code:
   `install.packages(<path_to_tar.gz_file>, repos = NULL, type = "source")`
6. Notes
    - Don\'t use Build (from top menu of RStudio) -> Install and Restart (Ctrl+Shift+B) because of
    problems explained [here](https://stackoverflow.com/questions/65004638/stop-in-package-function-doesnt-end-debug-mode).
        - [Here](https://stackoverflow.com/questions/1474081/how-do-i-install-an-r-package-from-source/1474125#1474125)
          can be found additional information e.g. regarding UNIX.
    - This package still doesn\'t have any stable version therefore previous installation steps
      are suggested.


### Try following steps if it seems that the package wasn\'t installed properly
1. Restart R session for each opened RStudio session
   - Continue with following steps if this doesn\'t help.
2. Uninstall this package from RStudio \"Packages\" tab if it exists there.
3. Delete directory `workflowrsubdirs` from directory `library`
   (directory with installed R packages) if it exists there.
4. Restart R session
   - for example using RStudio -> Session -> Restart R.
5. Install the package again using R code:
   install.packages(<path_to_tar.gz_file>, repos = NULL, type = \"source\").



## Necessary packages
It's necessary to install following packages for a proper usage
of this package (loading them isn't necessary):

- `base`
- `here`
- `rmarkdown`
- `stringr`
- `workflowr`
- `yaml`



## Examples
### Created solution
[.R files](https://github.com/LearnUseZone/rpkgs/tree/master/code/examples/iris),
their associated
[R Markdown files](https://github.com/LearnUseZone/rpkgs/tree/master/code-rmd/examples/iris)
and final [web pages](https://learnusezone.github.io/rpkgs/index.html)
show possible processing of iris dataset using package `workflowrsubdirs`. \
**The main focus of these examples is on demonstration of the functionality of package workflowrsubdirs rather than on choosing the best data interpretations.** \
These examples demonstrate for example:

- possible structure of directories
- correct functionality of `workflowrsubdirs::build_htmls()`
    - differences between `workflowrsubdirs::build_htmls()` and
      standard knitting (Knit button, `knitr::knit()`) are only:
        - `workflowrsubdirs::build_htmls()` is of course called instead of `knitr::knit()`
        - figures (e.g. when running graphics::hist()) are saved to directory `rpkgs/docs/figure`
        - directory structure of packages `workflowr` and `workflowrsubdirs` are used
          to keep data files organized
        - rules for using of package `workflowr` have to be met
- usage of a footer
    - File `footer.html` is saved in `rpkgs/analysis/include` which is
      in line with package `workflowr` (see e.g. [here](https://zenodo.org/record/3706876#.X9CkkLMo_cs)).
    - It's working in combination with definition of \"output\"
      in file `rpkgs/analysis/_site.yml` (commit 96087f9).
- usage of pictures
    - A picture `3-species-of-iris.png` saved
      in directory `rpkgs/docs/assets/examples/iris/static-visuals/histograms.Rmd`
      is used in `rpkgs/code-rmd/examples/iris/static-visuals/histograms.Rmd`.
    - A picture could be saved also in other part of `rpkgs/docs/assets` which is
      in line with package `workflowr` (read more
      [here](https://jdblischak.github.io/workflowr/articles/wflow-05-faq.html)).
- static visualizations with figures
    - File `rpkgs/code-rmd/examples/iris/static-visuals/histograms.Rmd` uses one .R file.
    - File `rpkgs/code-rmd/examples/iris/static-visuals/plots.Rmd` uses two .R files.
    - Both R Markdown files generate figures that are
      automatically (functionality of package `workflowr`) saved in directories
      `rpkgs/docs/figure/examples--iris--static-visuals--histograms.Rmd` and
      `rpkgs/docs/figure/examples--iris--static-visuals--plots.Rmd`.
    - Based on functionality of package `workflowr` if a .png file already exist,
      it's overwritten with a new file but remains saved even if it's deleted
      from the source R Markdown or .R file.
    - More information is
      [here](https://jdblischak.github.io/workflowr/articles/wflow-04-how-it-works.html#where-are-the-figures)
    - Also for example package ggplot2 is used (in \"boxplots.R\").
- interactive visualization
    - It's used in `rpkgs/code-rmd/examples/iris/interactive-visuals/plotly.Rmd`
      and `rpkgs/code-rmd/examples/iris/interactive-visuals/trelliscopejs.Rmd`.
    - It also uses:
      - [code externalization](https://bookdown.org/yihui/rmarkdown-cookbook/read-chunk.html)
      - packages: plotly, crosstalk, trelliscopejs
- usage of `knitr::opts_chunk$set()`
    - It's used in all R Markdown files.


### Brief example of steps from creating R codes until committing HTML pages
1. After `workflowrsubdirs` is installed, open a `workflowr` project (run .Rproj file).
2. Create .R files in directory `code` or its subdirectories.
3. Create R Markdown files in `code-rmd subdirs` (see above).
4. Optionally create other files like `footer.html` and update `_site.yml`.
5. Commit all previously created files.
   - Commit example: cb7ba9a on 21-02-09.
6. Run `workflowrsubdirs::build_htmls()` with relevant parameters to build HTML files
   that will be saved in directory `docs` (for GitHub) or `public` (for GitLab).
   - If you are sure that everything is set correctly (you don't need to test anymore),
     choose parameter `commit = TRUE`. It\'s necessary to commit in this step otherwise
     \"Checks\" of workflowr will show \"R Markdown file: uncommitted changes\".
   - Commit example: 9adb980 on 21-02-09.
   - Code example: `workflowrsubdirs::build_htmls("code-rmd/examples/iris", T, NULL, T)`
   - Note: commit d5cbbf4 is after running `workflowr::wflow_publish("analysis/*.Rmd")`.
7. This step can be done only if step 6 is committed (`workflowrsubdirs::build_htmls()` with `commit = TRUE`) \
   Remember that although temporary R Markdown files are committed separately when run function
   `workflowrsubdirs::build_htmls()`, it's still necessary to commit
   (run another code or use software like Sourcetree, \
   etc. = hereinafter \"manually commit\") all remaining
   (all deleted temporary R Markdown files and new HTML) files after HTML files are created.
   - Commit example: a0ae3fb on 21-02-09.


### Rendering options
Some of options to render R Markdown files to HTML files follow.

- Before using `workflowrsubdirs::build_htmls()`, remember to firstly run file `.Rproj`
  which also set the correct project working directory
  (read more in subtitle \"Open workflowr project before using workflowrsubdirs\").

```{r}
# render all .rmd and .Rmd files in directory "code-rmd" and its subdirectories to HTML files
#   note: all .rmd and .Rmd files anywhere under "code-rmd" will be processed
workflowrsubdirs::build_htmls()

# render all .rmd and .Rmd files in directory "code-rmd/examples" and its subdirectories to HTML files
#   note 1: all .rmd and .Rmd files anywhere under "code-rmd/examples" will be processed
#   note 2: all 3 following codes are equivalent
workflowrsubdirs::build_htmls(dir_path = "code-rmd/examples")
workflowrsubdirs::build_htmls(dir_path = "code-rmd\\examples")
workflowrsubdirs::build_htmls(dir_path = "code-rmd/examples", patterns = ".*.(r|R)md$")

# render .rmd and .Rmd files based on different regular expressions in parameter "patterns"
workflowrsubdirs::build_htmls(dir_path = "code-rmd/examples/iris",
                              subdirs = T,
                              patterns = c(
                                "-.*.[r,R]md",
                                "hist.*.Rmd$",
                                "plot.{1,2}\\.(?i)rmd",
                                "plotly.Rmd")
                              )
# the code I've actually used to generate all HTML files follows
#   note: previous examples generated HTML files but didn't create
#         separate commit of temporary analysis R Markdown files
#         (see input parameter commit) which means that after
#         a created HTML file is opened, workflowr button will show in "Checks" tab error:
#         R Markdown file: uncommitted changes
workflowrsubdirs::build_htmls("code-rmd/examples/iris", T, NULL, T)
```

- Notes
    1. Input parameter `dir_path` can also
       - contain the whole path to a directory,
       - starts with `./`, `.\\' and
       - contain some kind of typos like used a redundant `/` and `\\`
         at the beginning or/and end of a path.
    2. If also files in directory `analysis` need to be updated using package `workflowr`,
       a code like the following one can be used:
    
    ```{r}
    workflowr::wflow_publish(
        files = c("analysis/_site.yml", "analysis/about.Rmd",
                  "analysis/index.Rmd", "analysis/license.Rmd"),
        message = "feat(workflowrsubdirs): Update HTML examples using workflowr")
    ```


### Working with package trelliscopejs
Following is an example of steps currently working
for rendering a solution with package `trelliscopejs` to HTML:

1) Make the first 5 steps from \"Brief example of steps from creating R codes until committing HTML pages\".
   - Commit example: 41ca33c on 21-02-09.
   - If you need to re-create HTML file that uses package `trelliscopejs`
     as the 1st step delete relevant `trelliscopejs` directories.
2) Open the created .R file (that uses package `trelliscopejs`) ->
   source this file (click on icon \"Source\" = Source the contents of the active document)
    - For this purpose, parameter \"path\" specified
      inside function `trelliscopejs::facet_trelliscope()` in the .R file is used.
      - Example: path = \"trelliscopejs/regressions\".
    - New directory with `trelliscopejs` data is created directly in a relevant project working directory.
      - Example: \"D:\\Projects\\GitHub\\rpkgs\\trelliscopejs\".
3) Delete directory `trelliscopejs` in directory `docs` (for GitHub) or `public` (for GitLab)
   (inside a relevant project working directory) if directory `trelliscopejs` exists there. \
   Copy (don't move, but really copy) directory `trelliscopejs`,
   that was created by previous sourcing of .R code in point 2,
   from relevant project working directory to previously mentioned directory
   `docs` (for GitHub) or `public` (for GitLab).
4) Run `workflowrsubdirs::build_htmls()` with relevant arguments
   to build HTML files that will be saved in directory `docs` (for GitHub) or `public` (for GitLab).
   - If you are sure that everything is set correctly (you don\'t need to test anymore),
     choose parameter `commit = TRUE`. It\'s necessary to commit in this step otherwise
     \"Checks\" of workflowr will show \"R Markdown file: uncommitted changes\".
     - example if you want to build and commit all .Rmd files (some of them can contain `trelliscopejs`)
       anywhere in directory \"code-rmd\": \
       `workflowrsubdirs::build_htmls("code-rmd", T, NULL, T)`
   - If you run it with parameter `commit = TRUE` then a separate commit is automatically created, \
     commit example: 42c119a on 21-02-09.
5) Delete directory `trelliscopejs` from the relevant project working directory (as it was created in step 2).
    - Keep it in directory `docs` (for GitHub) or `public` (for GitLab).
6) This step can be done only if step 4 is committed (`workflowrsubdirs::build_htmls()` with `commit = TRUE`) \
   Manually commit all remaining `trelliscopejs` files,
   (automatically) deleted R Markdown file (from directory `analysis`) and added HTML file.
    - Commit example: ce8e20f on 21-02-09.
    - Note: If you build R Markdown file(s) containing `trelliscopejs` together with other
      R Markdown files, you need to commit more files.
7) Push your solution online. It may take few minutes to see working visualization online.



## How to avoid and solve potential problems
### YAML headers
- It's necessary to use correct quotation marks if inline R code is used in YAML header.
  otherwise there is an error with rendering R Markdown to HTML file.

- Example of using quotation marks (in this case: `"`) incorrectly:
  ```{r}
  date:    "`r paste("Last update:", format(lubridate::with_tz(as.POSIXct(Sys.time()) + 7200,
           tzone = "GMT"), "%Y-%m-%d %H:%M GMT+2"))`"
  ```

- Single quotation marks (\') have to be used in inline R codes in YAML header
  in R Markdown files saved in directory \"analysis\", for example: \
  ```{r}
  date:    "`r paste('Last update:', format(lubridate::with_tz(as.POSIXct(Sys.time()) + 7200,
           tzone = 'GMT'), '%Y-%m-%d %H:%M GMT+2'))`"
  ```

- Escaped double quotation marks (\\\") have to be used in inline R codes in YAML header
  in R Markdown files saved in \"code-rmd subdirs\", for example: \
  ```{r}
  date:    "`r paste(\"Last update:\", format(lubridate::with_tz(as.POSIXct(Sys.time()) + 7200,
           tzone = \"GMT\"), \"%Y-%m-%d %H:%M GMT+2\"))`"
  ```

- Examples of errors if incorrect quotation marks are used:
  - `workflowr::wflow_build("analysis/testfile.Rmd")`
    - `'`   -  working
    - `\'`  -  Scanner error: while parsing a quoted scalar at line 3, column 10 found unknown
               escape character at line 3, column 20
    - `"`   -  Parser error: while parsing a block mapping at line 1, column 1 did not find expected key at line 3, column 21
    - `\"`  -  Error: callr subprocess failed: \<text\>:1:7: unexpected input
    - `''`  -  Error: callr subprocess failed: \<text\>:1:9: unexpected symbol \
\
  - `workflowrsubdirs::build_htmls("code-rmd/subdir", F, "testfile.Rmd")`
    - `'`   -  Error: callr subprocess failed: \<text\>:1:9: unexpected symbol (\'\' is created)
    - `\'`  -  Scanner error: while parsing a quoted scalar at line 3, column 10 found unknown
               escape character at line 3, column 20
    - `"`   -  Parser error: while parsing a block mapping at line 1, column 1 did not find expected key at line 3, column 21
    - `\"`  -  working
    - `''`  -  Error: callr subprocess failed: \<text\>:1:9: unexpected string constant


### Problems when R Markdown file contains space in its name
- It's necessary to not use spaces in R Markdown file names otherwise following errors occur.
    - It's unusual to have file names, that will be used for web addresses, containing spaces
      so the easiest way how to avoid following errors is simply to not use spaces
      in R Markdown file names.
- Error when R Markdown file saved in `code-rmd subdirs` contain a space in its name after running
  e.g. following code is: Error: callr subprocess failed: cannot open the connection
    ```{r}
    workflowrsubdirs::build_htmls("code-rmd/subdir", F, "file.Rmd")
    ```
- Error when R Markdown file saved in directory `analysis` contain a space
  in its name after running e.g. following code is:
  Error: callr subprocess failed: The specified directory \'D:/Projects/GitLab/portfolios/test file.Rmd\' does not exist.
  (This error message is strange because directory \"analysis\" isn\'t written there.)
    ```{r}
    workflowr::wflow_build("analysis/file.Rmd")
    ```
- It seems that both of these errors are caused directly by package `workflowr`.
  J. Blischak (creator of package `workflowr`) will add a better error message
  (or maybe another suitable solution) as we discuss [here](https://github.com/jdblischak/workflowr/issues/226).


### Warning messages when using trelliscopejs::facet_trelliscope() with parameter as_plotly = TRUE
Running a solution that uses `trelliscopejs::facet_trelliscope()` with parameter `as_plotly = TRUE`
is working but a warning like the following one is shown (in Console).

```
writing thumbnail    [===============>-]  95% 39/41 eta: 1sPhantomJS not found. You can install it with webshot::install_phantomjs(). If it is installed, please make sure the phantomjs executable can be found via the PATH variable.
Warning message:                                           
In normalizePath(path.expand(path), winslash, mustWork) :
  path[1]="file://C:\Users\Master\AppData\Local\Temp\RtmpMRRyQJ\file43fc61851e76.html": The filename, directory name, or volume label syntax is incorrect
```

If you run `webshot::install_phantomjs()` and then run a relevant solution again
then warning like the following one is shown (in Console):

```
writing thumbnail    [==================>-]  95% 39/41 eta: 1sCould not load  file:///D:/Projects/GitHub/rpkgs/file:/C:/Users/Master/AppData/Local/Temp/RtmpwnUxl0/file1e6c48df84a.html
* could not create htmlwidget thumbnail... will use blank thumbnail...
Warning message:
  In normalizePath(path.expand(path), winslash, mustWork) :
  path[1]="file://C:\Users\Master\AppData\Local\Temp\RtmpwnUxl0\file1e6c48df84a.html": The filename, directory name, or volume label syntax is incorrect
```

This issue is also described [here](https://github.com/hafen/trelliscopejs/issues/61).



## Additional notes
- I'm still working on improvements (I don't have a stable version), some of them follow:
  - replace some placeholders like `tests\testthat\test-build_htmls.R` with a solution,
  - replace comments for future improvements with relevant improvements,
  - check and update comments and roxygen2 descriptions,
  - check and update some parts of R codes.
- This package `workflowrsubdirs` is available under the
  [MIT license](https://choosealicense.com/licenses/mit/).
- Initial inspiration for this package is from
  [this discussion](https://github.com/jdblischak/workflowr/issues/95). \
- A related discussion with John Blischak (a creator of package `workflowr`) about some of
  used concepts is [here](https://github.com/jdblischak/workflowr/issues/220).
